---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(warning = FALSE)
```

## Loading and preprocessing the data

The data is provided as a zip file so first we unzip it into a subdirectory called "data" and then load the csv file into a dataframe. The data as provided has the number of steps per five-minute interval so we need to calculate the number of steps per day by summing the intervals for each day.

```{r}
library(utils)
unzip('activity.zip', exdir = "data")
activity <- read.csv('data/activity.csv')

library(dplyr)

daily <- activity %>% 
        select(date, steps) %>%
        group_by(date) %>%
        summarize_each(funs(sum))

interval <- activity %>% 
        select(interval, steps) %>%
        group_by(interval) %>%
        summarize_each(funs(mean(., na.rm = TRUE)))
```



## What is mean total number of steps taken per day?

For this part of the assignment, you can ignore the missing values in
the dataset.

1. Make a histogram of the total number of steps taken each day

```{r}
library(ggplot2)
p <- ggplot(daily, aes(x=steps)) +
        geom_histogram()
p
```


2. Calculate and report the **mean** and **median** total number of steps taken per day

```{r}
meansteps <- mean(daily$steps, na.rm = TRUE)
print(paste("Mean number of steps per day = ", meansteps))
mediansteps <- median(daily$steps, na.rm = TRUE)
print(paste("Median number of steps per day = ", mediansteps))
```


## What is the average daily activity pattern?

1. Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis)
and the average number of steps taken, averaged across all days (y-axis)

```{r}
library(ggplot2)
p <- ggplot(interval, aes(x=interval, y=steps)) +
        geom_line()
p
```

2. Which 5-minute interval, on average across all the days in the dataset,
contains the maximum number of steps?

```{r}
maxinterval <- interval %>% filter(steps == max(steps))
maxinterval$interval
```



## Imputing missing values

Note that there are a number of days/intervals where there are missing values
(coded as NA). The presence of missing days may introduce bias into some
calculations or summaries of the data.
1. Calculate and report the total number of missing values in the dataset
(i.e. the total number of rows with NAs)

```{r}
missing <- activity %>% filter(is.na(steps))
count(missing)
```


2. Devise a strategy for filling in all of the missing values in the dataset. The
strategy does not need to be sophisticated. For example, you could use
the mean/median for that day, or the mean for that 5-minute interval, etc.

Use mean for the 5 minute interval

3. Create a new dataset that is equal to the original dataset but with the
missing data filled in.

```{r}
imputed <- replace(activity$steps, is.na(activity$steps), interval$steps)
activity_imputed <- activity
activity_imputed$steps<- imputed

daily_imputed <- activity_imputed %>% 
        select(date, steps) %>%
        group_by(date) %>%
        summarize_each(funs(sum))
        
```


4. Make a histogram of the total number of steps taken each day and Calculate
and report the mean and median total number of steps taken per day. Do
these values differ from the estimates from the first part of the assignment?
What is the impact of imputing missing data on the estimates of the total
daily number of steps?

```{r}
library(ggplot2)
p <- ggplot(daily_imputed, aes(x=steps)) +
        geom_histogram()
p
```

```{r}
meansteps_imputed <- mean(daily_imputed$steps, na.rm = FALSE)
print(paste("Mean number of steps per day = ", meansteps_imputed))
mediansteps_imputed <- median(daily_imputed$steps, na.rm = FALSE)
print(paste("Median number of steps per day = ", mediansteps_imputed))
```

The mean is unchanged, however replacing NA values with the respective means has changed the median, which is now equal to the overall mean.

## Are there differences in activity patterns between weekdays and weekends?

``` {r}
library(lubridate)
activity_imputed$date <- ymd(activity_imputed$date)
activity_imputed$day <- weekdays(activity_imputed$date)
activity_imputed$daytype <- ifelse((activity_imputed$day %in% c("Saturday", "Sunday")), "weekend", "weekday")

interval_imputed <- activity_imputed %>% 
        select(interval, steps, daytype) %>%
        group_by(interval, daytype) %>%
        summarize_each(funs(mean(., na.rm = TRUE)))
```


```{r}
library(ggplot2)
p <- ggplot(interval_imputed, aes(x=interval, y=steps)) +
        geom_line() +
        facet_grid(daytype~.)
p
```